---
title: "Analysing WHO STEPS Survey Data with stepssurvey"
author: "Abhijit Pakhare"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing WHO STEPS Survey Data with stepssurvey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  eval = TRUE
)
```

## Introduction

The WHO STEPwise Approach to NCD Risk Factor Surveillance (STEPS) is the
standard method for collecting population-level data on chronic disease
risk factors.  Surveys measure behavioural risk factors (tobacco, alcohol,
diet, physical activity), physical measurements (anthropometry, blood
pressure), and biochemical markers (blood glucose, cholesterol).

The **stepssurvey** package provides a complete, end-to-end analysis
pipeline that takes raw STEPS data from any country and produces
publication-ready tables, visualisations, and Word reports -- all while
properly accounting for the complex survey design (stratification,
clustering, sampling weights).

### What this guide covers

1. Installing the package
2. Understanding the data pipeline
3. Importing and detecting STEPS variables (v3.1 and v3.2)
4. Cleaning and deriving WHO-standard indicators
5. Setting up the complex survey design
6. Computing weighted prevalence estimates
7. Building tables and visualisations
8. Generating Word reports (fact sheet, data book, country report)
9. Running the full pipeline in one call
10. Working with your own data and customising the analysis


## Installation

Install the development version from GitHub:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("drpakhare/stepssurvey")
```

Or with **devtools**:

```{r install-devtools, eval = FALSE}
# install.packages("devtools")
devtools::install_github("drpakhare/stepssurvey")
```

Load the package:

```{r library}
library(stepssurvey)
```


## The pipeline at a glance

The package follows a linear, modular pipeline.  You can use the
one-command `run_steps_pipeline()` shortcut or call each step yourself
for full control.

```
  Raw data (.csv / .xlsx / .dta / .sav)
       |
       v
  import_steps_data()          -- read any format
       |
       v
  detect_steps_columns()       -- auto-detect v3.1/v3.2 codes
       |
       v
  clean_steps_data()           -- derive WHO indicators
       |
       v
  setup_survey_design()        -- step-specific weights
       |
       v
  compute_all_indicators()     -- weighted prevalences & means
       |
       +---> build_steps_tables()      -> summary flextables
       +---> build_steps_plots()       -> ggplot2 charts
       +---> render_country_report()   -> Summary Report (Word)
       |
       v
  compute_all_tables()         -- 60+ WHO registry tables
       |
       v
  build_all_tables()           -> 3-panel flextables (Men|Women|Both)
       |
       +---> render_data_book()        -> Detailed Data Book (Word)
```

Each function returns its output, so you can inspect, modify, or export
results at every stage.


## Step 1: Generate or import data

### Using the built-in test data generator

For learning and testing, the package includes a realistic data simulator:

```{r generate}
raw <- generate_test_data(n = 3000, seed = 42)
dim(raw)
names(raw)
```

The generated dataset mimics a real STEPS survey with 5 strata,
40 primary sampling units (PSUs), sampling weights, and realistic
correlations between risk factors (e.g. blood pressure increasing
with age, higher tobacco use in males).

### Importing real STEPS data

In practice you will have a data file exported from Epi Info, SPSS, or
Stata.  The `import_steps_data()` function reads all common formats and
standardises column names to lowercase with underscores:

```{r import, eval = FALSE}
# CSV
raw <- import_steps_data("data/raw/steps_survey_2024.csv")

# Excel
raw <- import_steps_data("data/raw/steps_survey_2024.xlsx")

# Stata (.dta) -- common for STEPS exports
raw <- import_steps_data("data/raw/steps_survey_2024.dta")

# SPSS (.sav)
raw <- import_steps_data("data/raw/steps_survey_2024.sav")
```

The function uses the file extension to pick the right reader
(`readr::read_csv`, `readxl::read_excel`, `haven::read_dta`, or
`haven::read_spss`), then passes column names through
`janitor::clean_names()` so that regardless of original casing you get
a consistent format like `wt_final`, `age`, `sex`.


## Step 2: Auto-detect STEPS variables

WHO STEPS datasets use standardised variable codes, but the codes changed
between instrument versions and many countries add their own prefixes.
The `detect_steps_columns()` function searches for each variable using a
prioritised alias list:

```{r detect}
cols <- detect_steps_columns(raw)
```

It returns a named list mapping each conceptual variable to the actual
column found in your data.  You can inspect the mapping:

```{r detect-inspect}
# Which column was matched for fasting glucose?
cols$fasting_glucose

# Which column for SBP reading 1?
cols$sbp1

# How many columns were detected?
sum(!sapply(cols, is.null))
```

### Version 3.1 vs 3.2 variable codes

A key feature of the package is transparent support for both WHO STEPS
instrument versions.  The variable codes changed substantially between
versions:

| Measurement        | v3.1 / Epi Info | v3.2 Instrument |
|--------------------|-----------------|-----------------|
| SBP readings       | B1, B3, B5      | M4a, M5a, M6a   |
| DBP readings       | B2, B4, B6      | M4b, M5b, M6b   |
| BP medications     | B7              | M7 / H3          |
| Height             | M1              | M11              |
| Weight             | M2              | M12              |
| Waist              | M3              | M14              |
| Fasting glucose    | C1 (c1_mmol)    | B5               |
| Diabetes meds      | C5              | B6 / H8          |
| Total cholesterol  | C6              | B8               |
| Cholesterol meds   | C10             | B9 / H14         |
| Sex                | --              | C1               |
| Age                | C1              | C3               |

The detection function includes aliases for both versions, so a dataset
using `b1` for SBP (v3.1) and one using `m4a` (v3.2) will both be
detected correctly.  The search is case-insensitive.

If a column is not found automatically, you can override the mapping
before cleaning:

```{r override, eval = FALSE}
cols$fasting_glucose <- "my_custom_glucose_variable"
```


## Step 3: Clean and derive indicators

The `clean_steps_data()` function performs all WHO-recommended data
processing in a single call:

```{r clean}
clean <- clean_steps_data(raw, cols, age_min = 18, age_max = 69)
dim(clean)
```

### What the cleaning step does

**Demographics:**

- Restricts age to the specified range (default 18--69)
- Creates WHO standard age groups: 18--24, 25--34, 35--44, 45--54,
  55--64, 65+
- Harmonises sex coding (1/2, "Male"/"Female", "M"/"F" all accepted)
- Ensures survey weight, stratum, and PSU columns are present

**Behavioural risk factors (Step 1):**

- Recodes tobacco and alcohol variables to logical TRUE/FALSE using
  `recode_yn()`, which understands 0/1, 1/2, "yes"/"no" patterns
- Computes average daily fruit and vegetable servings and flags
  `low_fruit_veg` (combined < 5 servings/day)
- Classifies physical activity into Low / Moderate / High based on
  MET-minutes/week thresholds (< 600, 600--2999, >= 3000)

**Physical measurements (Step 2):**

- Applies plausibility checks (e.g. height 100--250 cm, weight 20--300 kg,
  waist 40--200 cm) and sets implausible values to `NA`
- Computes BMI and classifies into Underweight / Normal / Overweight / Obese
- Flags central obesity using WHO waist circumference thresholds
  (>= 102 cm male, >= 88 cm female)
- Computes waist-to-hip ratio if both measurements are available
- Averages the last two of three BP readings (WHO protocol) to obtain
  mean SBP and mean DBP
- Creates the `raised_bp` indicator (SBP >= 140 or DBP >= 90 or on
  medication) and WHO blood pressure staging

**Biochemical measurements (Step 3):**

- Flags raised fasting glucose (>= 7.0 mmol/L or on diabetes medication)
- Flags impaired fasting glucose (6.1--6.9 mmol/L)
- Flags raised total cholesterol (>= 5.0 mmol/L)
- Flags low HDL cholesterol (sex-specific thresholds)
- Flags raised triglycerides (>= 1.7 mmol/L)

You can inspect the derived variables:

```{r clean-inspect}
# BMI categories
table(clean$bmi_category, clean$sex)

# Blood pressure staging
table(clean$bp_stage)

# Physical activity levels
table(clean$pa_category, clean$sex)
```


## Step 4: Set up the survey design

STEPS surveys use complex sampling designs.  Ignoring the design leads to
biased estimates and incorrect confidence intervals.  The
`setup_survey_design()` function creates a `survey::svydesign` object
that accounts for weights, stratification, and clustering:

```{r design}
designs <- setup_survey_design(clean)
```

The returned object is a list with three elements (`step1`, `step2`,
`step3`), each a `survey::svydesign` object weighted appropriately
for that step of the survey.  Functions like `compute_all_indicators()`
accept this list directly, but for custom estimates you pick the
design matching the step of the variable you are analysing:

The function auto-detects the design complexity based on which columns
are present:

- **Full complex design**: weights + strata + clusters
- **Weights + clusters**: no stratification variable
- **Weights + strata**: no clustering (rare)
- **Weights only**: self-representing design
- **Unweighted**: simple random sample (weights set to 1)

Extreme sampling weights are automatically trimmed to the range
\[0.1, 10\] using `survey::trimWeights()` to prevent a few observations
from dominating the estimates.

The returned object can be used with any function from the **survey**
package if you need custom analyses beyond what the package provides.


## Step 5: Compute indicators

### All indicators at once

```{r indicators}
result <- compute_all_indicators(designs)
```

This returns a list with two elements:

- `result$results` -- a nested list of domain-specific estimates
  (total, by sex, by age group)
- `result$key_indicators` -- a tidy data frame of headline prevalences

```{r key}
result$key_indicators
```

### Domain-specific functions

For more control, call each domain function separately:

```{r domain, eval = FALSE}
tob  <- compute_tobacco_indicators(designs$step1)
alc  <- compute_alcohol_indicators(designs$step1)
diet <- compute_diet_pa_indicators(designs$step1)
anth <- compute_anthropometry_indicators(designs$step2)
bp   <- compute_bp_indicators(designs$step2)
bio  <- compute_biochemical_indicators(designs$step3)
```

Each returns a named list.  For example, the tobacco module returns:

```{r tob-example}
tob <- compute_tobacco_indicators(designs$step1)
names(tob)

# Overall prevalence of current tobacco use
tob$current_tobacco_total

# Prevalence by sex
tob$current_tobacco_by_sex
```

### Custom weighted estimates

The package exports two low-level helpers for any weighted estimate you
need:

```{r custom}
# Weighted proportion with 95% CI (raised_bp is a Step 2 variable)
svyprop(~raised_bp, designs$step2)

# Stratified by sex
svyprop(~raised_bp, designs$step2, by = ~sex)

# Weighted mean with 95% CI
svymn(~mean_sbp, designs$step2, by = ~sex)
```


## Step 6: Build publication-ready tables

The package provides **two table systems** for different purposes.

### Summary tables (Both Sexes only)

```{r tables}
tables <- build_steps_tables(result$results)
names(tables)
```

Each table is a **flextable** object styled with WHO STEPS branding
(dark blue headers, formatted confidence intervals).  These tables show
estimates by age group for Both Sexes combined -- ideal for summary
reports and quick reference.

```{r table-show, eval = FALSE}
# Display the raised blood pressure table
tables$raised_bp

# Export to Word
flextable::save_as_docx(tables$raised_bp, path = "bp_table.docx")
```

### Detailed WHO 3-panel tables (Men | Women | Both Sexes)

For the full WHO STEPS data book format, use the detailed table engine.
This produces ~60 tables in the standard 3-panel layout (Age Group | Men |
Women | Both Sexes):

```{r detailed-tables, eval = FALSE}
# Step 1: Compute raw results from the table registry
computed <- compute_all_tables(designs)

# Step 2: Format into flextable objects with WHO styling
detailed <- build_all_tables(computed)
names(detailed)  # e.g. "T_current_smokers", "M_bp_mean", "B_glucose_raised"
```

The table IDs use prefixes matching WHO STEPS domains:

| Prefix | Domain                    |
|--------|---------------------------|
| T_     | Tobacco                   |
| A_     | Alcohol                   |
| D_     | Diet                      |
| P_     | Physical Activity         |
| H_     | Health History & Treatment|
| M_     | Physical Measurements     |
| B_     | Biochemical Measurements  |
| R_     | Cardiovascular Risk       |
| RF_    | Combined Risk Factors     |

You can access individual tables by ID or filter by section:

```{r registry, eval = FALSE}
# Browse the full table registry
registry <- steps_table_registry()

# Get all tables for a specific section
bp_entries <- get_registry_by_section("Blood Pressure")

# Get all Step 2 tables
step2_entries <- get_registry_by_step(2)

# List available sections
list_registry_sections()
```


## Step 7: Create visualisations

```{r plots, fig.width = 8, fig.height = 5}
plots <- build_steps_plots(
  indicators    = result$results,
  key_indicators = result$key_indicators,
  country_name  = "Exampleland",
  survey_year   = 2024
)
names(plots)
```

### Overview chart

The overview plot shows all key indicators as a horizontal bar chart with
95% confidence intervals, sorted by prevalence:

```{r overview, fig.width = 8, fig.height = 5}
plots$overview
```

### Sex-stratified dashboard

If multiple sex-stratified indicators are available, the package creates
a 2 x 2 dashboard using **patchwork**:

```{r dashboard, fig.width = 10, fig.height = 7, eval = FALSE}
plots$sex_dashboard
```

### Age-stratified trends

Age trend plots show how each risk factor varies across the WHO standard
age groups, with shaded confidence bands:

```{r age-trend, fig.width = 8, fig.height = 4.5}
plots$bp_by_age
```

### Saving plots

```{r save-plots, eval = FALSE}
save_steps_plots(plots, output_dir = "outputs/figures")
# Creates:
#   outputs/figures/01_overview_indicators.png
#   outputs/figures/02_by_sex_dashboard.png
#   outputs/figures/03_bp_by_age.png
#   outputs/figures/04_obesity_by_age.png
```

### WHO STEPS colour palette and theme

The package uses a consistent visual identity.  You can apply the same
styling to your own ggplot2 plots:

```{r theme, fig.width = 6, fig.height = 3.5}
pal <- steps_colors()
str(pal)

library(ggplot2)
ggplot(clean, aes(x = age_group, fill = sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c(Male = pal$male, Female = pal$female)) +
  theme_steps() +
  labs(title = "Sample distribution by age and sex")
```


## Step 8: Generate Word reports

The package produces two complementary reports:

| Report | Function | Content |
|--------|----------|---------|
| **Summary Report** | `render_country_report()` | Narrative with key findings, charts, and recommendations |
| **Detailed Data Book** | `render_data_book()` | Complete WHO 3-panel tables (Men \| Women \| Both Sexes) across all domains |

```{r reports, eval = FALSE}
cfg <- steps_config(
  data_path    = "data/raw/steps_survey_2024.csv",
  country_name = "Exampleland",
  survey_year  = 2024,
  age_min      = 18,
  age_max      = 69
)

# Summary report -- narrative with key findings, charts, recommendations
render_country_report(cfg, output_dir = "outputs")

# Data book -- detailed WHO 3-panel tables by domain
render_data_book(cfg, output_dir = "outputs")

# Fact sheet -- one-page overview of key indicators
render_fact_sheet(cfg, output_dir = "outputs")
```

Each function runs the entire pipeline internally (import, clean, analyse)
and renders an R Markdown template to a Word document.  The output files
are saved in the specified directory.

### What each report contains

The **Summary Report** includes an executive summary table, narrative
sections for each risk factor domain with inline prevalence figures,
embedded charts (overview indicators, by-sex breakdowns, age trends),
and WHO-aligned policy recommendations.

The **Data Book** contains the full set of ~60 WHO STEPS tables in the
standard 3-panel format.  Each table shows estimates by age group
separately for Men, Women, and Both Sexes.  Tables are organised by
STEPS step: Step 1 (Behavioural), Step 1.5 (Health History), Step 2
(Physical Measurements), Step 3 (Biochemical), and Combined Risk Factors.


## One-command pipeline

For the fastest path from raw data to results, `run_steps_pipeline()`
chains every step and returns all intermediate objects:

```{r pipeline, eval = FALSE}
out <- run_steps_pipeline(
  data_path      = "data/raw/steps_survey_2024.csv",
  country_name   = "Exampleland",
  survey_year    = 2024,
  age_min        = 18,
  age_max        = 69,
  output_dir     = "outputs",
  render_reports = TRUE
)

# Access any intermediate result
out$raw_data
out$clean_data
out$design
out$indicators
out$key_indicators
out$tables
out$plots
```

Setting `render_reports = FALSE` skips the Word documents (useful for
interactive exploration or when **rmarkdown** / Pandoc are not available).


## Working with real STEPS data

### Preparing your data file

The package accepts data in four formats:

| Format     | Extension | Typical source                   |
|------------|-----------|----------------------------------|
| CSV        | .csv      | Spreadsheet export               |
| Excel      | .xlsx     | Direct data entry                |
| Stata      | .dta      | WHO Epi Info / analysis template |
| SPSS       | .sav      | SPSS data export                 |

Before importing, ensure the file contains at minimum:

- **Age** and **sex** columns (required for all analyses)
- **Sampling weight** column (recommended; if absent, all weights are
  set to 1)
- At least some risk factor measurements from Step 1, 2, or 3

### Handling custom variable names

If your dataset uses non-standard column names that are not auto-detected,
override the mapping after detection:

```{r custom-names, eval = FALSE}
raw  <- import_steps_data("my_steps_data.csv")
cols <- detect_steps_columns(raw)

# Suppose your glucose column is named "blood_sugar_fasting"
cols$fasting_glucose <- "blood_sugar_fasting"

# Suppose BP columns use a different naming pattern
cols$sbp1 <- "systolic_bp_1"
cols$dbp1 <- "diastolic_bp_1"

# Then proceed as normal
clean   <- clean_steps_data(raw, cols)
designs <- setup_survey_design(clean)
result  <- compute_all_indicators(designs)
```

### Adjusting age range

Some STEPS surveys target populations outside the standard 18--69 range.
Adjust with the `age_min` and `age_max` parameters:

```{r age-range, eval = FALSE}
# Include up to age 79
clean <- clean_steps_data(raw, cols, age_min = 18, age_max = 79)
```

Note that changing the upper age limit adds a wider final age group
(e.g. 65--79 instead of 65+).


## Complete worked example

This section walks through a full analysis using simulated data, showing
every step from generation to output.

```{r worked, fig.width = 8, fig.height = 5}
library(stepssurvey)

# 1. Generate a realistic test dataset
raw <- generate_test_data(n = 3000, seed = 42)

# 2. Detect standard STEPS variable columns
cols <- detect_steps_columns(raw)

# 3. Clean data and derive all indicators
clean <- clean_steps_data(raw, cols, age_min = 18, age_max = 69)

# 4. Create the complex survey design
designs <- setup_survey_design(clean)

# 5. Compute all NCD risk factor indicators
result <- compute_all_indicators(designs)

# 6. View headline estimates
result$key_indicators

# 7. Build formatted tables
tables <- build_steps_tables(result$results)

# 8. Build visualisations
plots <- build_steps_plots(
  indicators     = result$results,
  key_indicators = result$key_indicators,
  country_name   = "Exampleland",
  survey_year    = 2024
)

# 9. Display the overview chart
plots$overview
```


## WHO standard definitions used

The package implements the following WHO STEPS definitions for all
derived indicators:

| Indicator                     | Definition                                                                  |
|-------------------------------|-----------------------------------------------------------------------------|
| Current tobacco use           | Currently smokes any tobacco product (T1 = Yes)                             |
| Daily tobacco use             | Smokes tobacco daily (T2 = Yes)                                             |
| Current alcohol use           | Consumed alcohol in the past 30 days (A5 = Yes)                             |
| Heavy episodic drinking       | 6 or more standard drinks on a single occasion in past 30 days (A9)         |
| Insufficient physical activity| Total MET-minutes per week < 600                                            |
| Low fruit and vegetable intake| Combined < 5 servings per day                                               |
| Overweight or obese           | BMI >= 25 kg/m^2^                                                           |
| Obese                         | BMI >= 30 kg/m^2^                                                           |
| Central obesity               | Waist >= 102 cm (male) or >= 88 cm (female)                                |
| Raised blood pressure         | Mean SBP >= 140 mmHg or mean DBP >= 90 mmHg or currently on BP medication   |
| Raised fasting glucose        | Fasting blood glucose >= 7.0 mmol/L or currently on diabetes medication     |
| Impaired fasting glucose      | Fasting blood glucose 6.1--6.9 mmol/L                                       |
| Raised total cholesterol      | Total cholesterol >= 5.0 mmol/L                                             |
| Low HDL cholesterol           | HDL < 1.0 mmol/L (male) or < 1.3 mmol/L (female)                           |
| Raised triglycerides          | Triglycerides >= 1.7 mmol/L                                                 |

Blood pressure readings follow the WHO protocol of averaging the last two
of three measurements taken three minutes apart.


## FAQ

**Can I use this package with STEPS surveys from any country?**
Yes.
The variable detection system supports both v3.1 and v3.2 naming
conventions, plus common country-specific aliases.  Override any
undetected columns manually as shown above.

**What if my dataset is missing some risk factor modules?**
The package handles missing modules gracefully.  If, for example, no
biochemical columns are found, the glucose and cholesterol indicators
are simply skipped and the tables and plots adapt accordingly.

**Can I add my own indicators?**
Absolutely.  After the cleaning step you have a standard data frame with
all derived variables.  Use the `survey::svydesign` object with `svyprop()`
or `svymn()` (or any **survey** package function) for custom analyses.

**How do I cite this package?**
```{r cite, eval = FALSE}
citation("stepssurvey")
```


## Further resources

- [WHO STEPS Manual](https://www.who.int/teams/noncommunicable-diseases/surveillance/systems-tools/steps/manuals)
- [STEPS Instrument v3.2 (Q-by-Q Guide)](https://www.who.int/teams/noncommunicable-diseases/surveillance/systems-tools/steps/instrument)
- [STEPS Data Analysis Tools](https://www.who.int/teams/noncommunicable-diseases/surveillance/systems-tools/steps/data-analysis-reporting-tools)
- [Package source on GitHub](https://github.com/drpakhare/stepssurvey)


## Session info

```{r session}
sessionInfo()
```
