---
title: "Validation Against WHO Official Fact Sheets"
author: "Abhijit Pakhare"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation Against WHO Official Fact Sheets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Acknowledgements

This validation relies on several resources developed and maintained by
the World Health Organization (WHO):

- **WHO STEPS instrument and methodology**: The STEPwise Approach to NCD
  Risk Factor Surveillance (STEPS) was developed by the WHO Department
  of Noncommunicable Diseases. The instrument, survey manuals, and
  analysis guidelines (including the *STEPS Part 4: Guide to Physical
  Measurements* and *Part 5: Data Management and Analysis*) define the
  standard definitions and indicator computations used throughout this
  package.

- **WHO official STEPS analysis scripts**: The R scripts maintained by
  WHO for STEPS data processing and databook generation
  (available at <https://github.com/WorldHealthOrganization/stepsKGZ>)
  served as the authoritative reference for indicator derivation logic,
  survey design setup, weight handling, and data quality filters
  (`smk_cln`, `smkless_cln`). Key methodological decisions in
  **stepssurvey** — such as not trimming sampling weights, setting
  skip-pattern NAs to zero for denominator alignment, and applying
  screening-question logic for GPAQ — were directly informed by
  studying these scripts.

- **WHO STEPS country fact sheets**: Published fact sheet values for
  Moldova (2021), Mongolia (2019), and Georgia (2016) were used as the
  gold-standard reference estimates. These fact sheets are publicly
  available from the WHO NCD Microdata Repository
  (<https://extranet.who.int/ncdsmicrodata>).

- **Licensed STEPS microdata**: The underlying survey microdata were
  accessed under license from the WHO NCD Microdata Repository. Use of
  these data is subject to the WHO STEPS data access agreement, and the
  microdata files are not redistributable.

We gratefully acknowledge the national STEPS survey teams and
coordinators in the Republic of Moldova, Mongolia, and Georgia for
designing, conducting, and managing these surveys and for making the
data available for research use through the WHO NCD Microdata
Repository. The quality and completeness of these datasets made
systematic validation possible.

We also acknowledge the WHO NCD Surveillance team for developing and
maintaining the STEPS methodology, analysis tools, and data sharing
infrastructure that underpin this work.

## Overview

The **stepssurvey** package has been validated against WHO-published fact
sheet values for three national STEPS surveys spanning different regions,
time periods, and STEPS instrument versions:

| Country               | Year | Age range | Instrument | N (unweighted) |
|:----------------------|:----:|:---------:|:----------:|:--------------:|
| Republic of Moldova   | 2021 | 18--69    | v3.2       | 4,097          |
| Mongolia              | 2019 | 15--69    | v3.2       | 6,647          |
| Georgia               | 2016 | 18--69    | v3.1       | 5,765          |

For each country, the package pipeline (import → detect → clean → survey
design → indicator computation) was run end-to-end and results compared
with the "Both Sexes" estimates published in the corresponding WHO STEPS
country fact sheet.

## Validation criteria

Two complementary criteria were used to judge agreement:

1. **Point-estimate match** — the package estimate falls within 1.0
   percentage point (pp) of the fact sheet value for proportions, or 1.0
   unit for continuous measures (BMI, SBP, cholesterol).
2. **Confidence-interval overlap** — the 95 % confidence intervals from
   the package and the fact sheet share at least one common value.

The 1 pp tolerance accounts for rounding at different stages and minor
differences in the treatment of edge cases (e.g. "don't know" responses
coded as 77 or 88).

## Indicators compared

Up to 17 indicators were compared per country, covering all three STEPS
Steps:

**Step 1 — Behavioural risk factors:**
current tobacco use, current smoking, second-hand smoke exposure (home
and workplace), current alcohol use (past 30 days), heavy episodic
drinking, insufficient fruit and vegetable consumption, insufficient
physical activity.

**Step 2 — Physical measurements:**
mean BMI, overweight (BMI ≥ 25), obesity (BMI ≥ 30), mean systolic
blood pressure, raised blood pressure or on medication.

**Step 3 — Biochemical measurements:**
mean total cholesterol, raised cholesterol or on medication, raised
fasting glucose or on medication, impaired fasting glucose.

## Results summary

```{r summary-table, eval=TRUE, echo=FALSE, results='asis'}
summary_data <- data.frame(
  Country = c("Moldova 2021", "Mongolia 2019", "Georgia 2016", "**TOTAL**"),
  Compared = c(17, 15, 15, 47),
  Within_1pp = c(16, 15, 15, 46),
  CI_Overlap = c(17, 15, 15, 47),
  Match_Rate = c("94%", "100%", "100%", "**98%**")
)

knitr::kable(
  summary_data,
  col.names = c("Country", "Indicators compared", "Within 1 pp",
                "CI overlap", "Match rate"),
  align = c("l", "c", "c", "c", "c"),
  caption = "Validation summary: stepssurvey vs. WHO fact sheets"
)
```

Overall, **46 of 47 indicators (98 %) match within 1 pp**, and
**all 47 (100 %) have overlapping confidence intervals**.

## Detailed results by country

### Moldova 2021

```{r moldova, eval=TRUE, echo=FALSE, results='asis'}
mda <- data.frame(
  Indicator = c(
    "Current tobacco use", "Current smoking", "Second-hand smoke (home)",
    "Second-hand smoke (work)", "Current alcohol (30 d)",
    "Heavy episodic drinking", "Insufficient fruit/veg",
    "Insufficient physical activity",
    "Overweight (BMI >= 25)", "Obese (BMI >= 30)",
    "Raised BP or on meds",
    "Raised glucose or on meds", "Raised cholesterol or on meds",
    "Impaired fasting glucose",
    "Mean BMI", "Mean SBP", "Mean total cholesterol"
  ),
  Ours = c(27.7, 27.6, 23.3, 26.4, 63.2, 13.8, 63.2, 9.2,
           63.9, 22.7, 35.0, 6.3, 28.4, 9.9, 26.9, 129.2, 4.4),
  WHO = c(29.9, 27.6, 23.2, 26.4, 63.2, 13.8, 63.4, 9.1,
          63.9, 22.7, 34.8, 6.3, 27.7, 9.9, 26.9, 129.2, 4.4),
  Diff = c(-2.2, 0.0, 0.1, 0.0, 0.0, 0.0, -0.2, 0.1,
           0.0, 0.0, 0.2, 0.0, 0.7, 0.0, 0.0, 0.0, 0.0),
  Match = c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes",
            "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes"),
  CI_Overlap = c("Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes",
                 "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes")
)

knitr::kable(mda, align = c("l", "r", "r", "r", "c", "c"),
             caption = "Moldova 2021: 16/17 within 1 pp, 17/17 CI overlap")
```

The single mismatch is **current tobacco use** (−2.2 pp). The confidence
intervals overlap (ours: 25.6–29.8 vs. WHO: 27.7–32.1), suggesting the
difference may reflect rounding or a slightly different indicator
definition in the published fact sheet.

### Mongolia 2019

```{r mongolia, eval=TRUE, echo=FALSE, results='asis'}
mng <- data.frame(
  Indicator = c(
    "Current tobacco use", "Current smoking (daily)",
    "Current alcohol (30 d)",
    "Heavy episodic drinking", "Insufficient fruit/veg",
    "Insufficient physical activity",
    "Overweight (BMI >= 25)", "Obese (BMI >= 30)",
    "Raised BP or on meds (130/80)",
    "Raised glucose or on meds", "Raised cholesterol or on meds",
    "Impaired fasting glucose",
    "Mean BMI", "Mean SBP", "Mean total cholesterol"
  ),
  Ours = c(25.0, 21.6, 34.8, 20.2, 83.2, 22.5,
           49.3, 18.5, 44.3, 8.3, 27.8, 17.4, 25.6, 120.5, 4.4),
  WHO = c(24.2, 21.6, 34.8, 19.8, 83.4, 21.9,
          49.4, 18.5, 44.0, 8.3, 27.8, 17.4, 25.5, 120.5, 4.4),
  Diff = c(0.8, 0.0, 0.0, 0.4, -0.2, 0.6,
           -0.1, 0.0, 0.3, 0.0, 0.0, 0.0, 0.1, 0.0, 0.0),
  Match = rep("Yes", 15),
  CI_Overlap = rep("Yes", 15)
)

knitr::kable(mng, align = c("l", "r", "r", "r", "c", "c"),
             caption = "Mongolia 2019: 15/15 within 1 pp, 15/15 CI overlap")
```

Mongolia uses a raised blood-pressure threshold of **130/80 mmHg**
(rather than the standard 140/90), and the age range begins at 15
(rather than 18). Second-hand smoke indicators were not reported in the
Mongolia fact sheet.

Note: the Mongolia WHO fact sheet reports **daily smoking** under the
"current smoking" label. The package computes both any-current and
daily smoking; the daily-smoking variable is used for this comparison.

### Georgia 2016

```{r georgia, eval=TRUE, echo=FALSE, results='asis'}
geo <- data.frame(
  Indicator = c(
    "Current tobacco use", "Current smoking (daily)",
    "Current alcohol (30 d)",
    "Heavy episodic drinking", "Insufficient fruit/veg",
    "Insufficient physical activity",
    "Overweight (BMI >= 25)", "Obese (BMI >= 30)",
    "Raised BP or on meds",
    "Raised glucose or on meds", "Raised cholesterol or on meds",
    "Impaired fasting glucose",
    "Mean BMI", "Mean SBP", "Mean total cholesterol"
  ),
  Ours = c(31.1, 28.0, 39.0, 18.7, 62.9, 18.2,
           64.6, 33.4, 37.7, 4.5, 27.7, 2.0, 28.2, 129.4, 4.3),
  WHO = c(31.0, 28.0, 39.1, 18.3, 63.0, 17.4,
          64.6, 33.2, 37.7, 4.5, 27.7, 2.0, 28.1, 129.4, 4.3),
  Diff = c(0.1, 0.0, -0.1, 0.4, -0.1, 0.8,
           0.0, 0.2, 0.0, 0.0, 0.0, 0.0, 0.1, 0.0, 0.0),
  Match = rep("Yes", 15),
  CI_Overlap = rep("Yes", 15)
)

knitr::kable(geo, align = c("l", "r", "r", "r", "c", "c"),
             caption = "Georgia 2016: 15/15 within 1 pp, 15/15 CI overlap")
```

As with Mongolia, the Georgia WHO fact sheet reports **daily smoking**
under the "current smoking" label. Second-hand smoke indicators were
not reported in the Georgia fact sheet.

## Key methodological alignments

During validation, several methodological details were identified where
alignment with the WHO official analysis scripts was essential for
reproducibility:

**Survey design.** Sampling weights are used as-is without trimming,
matching the WHO official STEPS analysis scripts. Lone PSUs are handled
with `options(survey.lonely.psu = "adjust")`.

**Alcohol skip patterns.** Non-drinkers who skip the heavy-episodic
drinking question (A9) are coded as `FALSE` (not `NA`), ensuring the
denominator covers the total population rather than only current
drinkers.

**Physical activity screening questions.** GPAQ screening questions
(P1, P4, P7, P10, P13) where the respondent answers "No" set the
domain contribution to zero, rather than leaving it as `NA`. Special
codes 77 and 88 ("don't know" / "refused") are cleaned to `NA`.

**Diet zero-days handling.** When a respondent reports eating fruit or
vegetables on zero days per week, the servings-per-day variable (which
is skipped and therefore `NA`) is set to zero rather than excluded from
the denominator.

**Configurable blood-pressure thresholds.** The package supports custom
SBP/DBP thresholds via `clean_steps_data(bp_sbp_threshold, bp_dbp_threshold)`,
needed because some countries (e.g. Mongolia) use thresholds other than
the standard 140/90.

**Data quality filters.** The WHO `smk_cln` filter is applied: logically
inconsistent tobacco respondents (e.g. "I currently smoke" combined with
"I never smoked in the past") are excluded from the tobacco denominator.

**Column mapping.** Country datasets often use non-standard variable
names.  The package's `read_column_mapping()` function and Excel mapping
template were used to handle datasets where auto-detection required
manual overrides.

## Running the validation yourself

The full validation script is included in the package at
`inst/validation/validate_all.R`. To run it, you need the licensed STEPS
microdata files placed in a `STEPS Licensed Datasets` directory alongside
the package:

```r
library(stepssurvey)
setwd("path/to/stepssurvey")
source("inst/validation/validate_all.R")
```

Per-country results are saved as CSV files in `inst/validation/`.

## Reproducibility notes

The validation was performed with R `r paste0(R.version$major, ".", R.version$minor)`,
using survey package version `r packageVersion("survey")`, and
stepssurvey version `r packageVersion("stepssurvey")`.

The licensed STEPS microdata files are not redistributable and are
therefore not included in the package. Researchers with access to the
WHO STEPS microdata repository can request datasets for the three
countries listed above and reproduce these results independently.

## References

1. World Health Organization. (2017). *WHO STEPS Surveillance Manual:
   The WHO STEPwise Approach to Noncommunicable Disease Risk Factor
   Surveillance.* Geneva: WHO.
   <https://www.who.int/teams/noncommunicable-diseases/surveillance/systems-tools/steps>

2. World Health Organization. *WHO NCD Microdata Repository: STEPS
   Survey Data.* <https://extranet.who.int/ncdsmicrodata>

3. World Health Organization. *STEPS data analysis scripts (R).*
   GitHub repository.
   <https://github.com/WorldHealthOrganization/stepsKGZ>

4. Republic of Moldova STEPS Survey 2021. *WHO NCD Country Fact
   Sheet.* Geneva: WHO, 2022.

5. Mongolia STEPS Survey 2019. *WHO NCD Country Fact Sheet.* Geneva:
   WHO, 2020.

6. Georgia STEPS Survey 2016. *WHO NCD Country Fact Sheet.* Geneva:
   WHO, 2017.

7. Lumley, T. (2004). Analysis of complex survey samples. *Journal of
   Statistical Software*, 9(8), 1--19. R package **survey** version
   `r packageVersion("survey")`.
