---
title: "NCD Risk Factor Survey Report"
subtitle: "`r params$config$country_name` | `r params$config$survey_year`"
output:
  word_document: default
  html_document:
    theme: flatly
    toc: true
params:
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10)
library(flextable); library(ggplot2); library(glue); library(patchwork)
library(stepssurvey)

cfg            <- params$config
indicators     <- readRDS(file.path(cfg$data_path, "..", "processed", "indicators.rds"))
key_indicators <- readRDS(file.path(cfg$data_path, "..", "processed", "key_indicators.rds"))
tables         <- readRDS(file.path(cfg$data_path, "..", "processed", "tables.rds"))
plots          <- readRDS(file.path(cfg$data_path, "..", "processed", "plots.rds"))

# Helper to pull a single estimate
get_est <- function(ind, domain, key) {
  val <- ind[[domain]][[key]]
  if (is.null(val)) return("N/A")
  paste0(round(val$estimate[1], 1), "%")
}
```

# Executive Summary

This report presents the findings of the WHO STEPwise Approach to NCD Risk Factor Surveillance (STEPS) survey conducted in **`r cfg$country_name`** in **`r cfg$survey_year`**. The survey collected data on key non-communicable disease (NCD) risk factors among adults aged `r cfg$age_min`–`r cfg$age_max` years.

## Key Findings

```{r key-findings}
key_indicators |>
  mutate(Result = paste0(round(estimate, 1), "% (95% CI: ", round(lower, 1), "–", round(upper, 1), "%)")) |>
  select(Domain = domain, Indicator = indicator, Result) |>
  flextable() |>
  theme_vanilla() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#00427A") |>
  color(part = "header", color = "white") |>
  merge_v(j = "Domain") |>
  autofit()
```

---

# Tobacco Use

Current tobacco use was estimated at **`r get_est(indicators, "tobacco", "current_tobacco_total")`** among the adult population. Tobacco use is a leading risk factor for cardiovascular disease, cancer, and chronic respiratory disease.

```{r tobacco-chart, fig.height=5}
if (!is.null(plots$tobacco_by_sex)) plots$tobacco_by_sex
```

---

# Physical Activity

Insufficient physical activity (less than 600 MET-minutes/week) was found in **`r get_est(indicators, "diet_pa", "insufficient_pa_total")`** of the population. Regular physical activity reduces the risk of cardiovascular disease, diabetes, and some cancers.

---

# Overweight and Obesity

The prevalence of overweight or obesity (BMI ≥25 kg/m²) was **`r get_est(indicators, "anthropometry", "overweight_obese_total")`**.

```{r obesity-age, fig.height=5}
if (!is.null(plots$obesity_by_age)) plots$obesity_by_age
```

---

# Blood Pressure

Raised blood pressure (≥140/90 mmHg or on antihypertensive medication) was found in **`r get_est(indicators, "blood_pressure", "raised_bp_total")`** of adults.

```{r bp-age, fig.height=5}
if (!is.null(plots$bp_by_age)) plots$bp_by_age
```

---

# Blood Glucose and Cholesterol

Raised fasting blood glucose (≥7.0 mmol/L or on diabetes medication) was found in **`r get_est(indicators, "biochemical", "raised_glucose_total")`** of adults. Raised total cholesterol (≥5.0 mmol/L) was found in **`r get_est(indicators, "biochemical", "raised_chol_total")`**.

---

# Recommendations

Based on the survey findings, the following public health priorities are recommended:

1. **Strengthen tobacco control** policies including taxation, plain packaging, and cessation support
2. **Promote physical activity** through community programs and urban planning
3. **Improve dietary quality** by increasing access to fruits and vegetables
4. **Scale up screening and treatment** for hypertension and diabetes
5. **Enhance surveillance capacity** with regular repeat surveys

---

# Methodology

This survey followed the WHO STEPwise approach to NCD risk factor surveillance. Data were collected using a multistage cluster sampling design. Survey-weighted estimates were produced using the `survey` package in R to account for the complex sampling design. Confidence intervals were calculated at the 95% level.

---

*Report generated: `r Sys.Date()`*
*Analysis: WHO STEPS R Pipeline*
*Country: `r cfg$country_name` | Survey year: `r cfg$survey_year`*
