---
title: "STEPS Survey Data Book"
subtitle: "NCD Risk Factor Survey — Detailed Tables"
output:
  word_document: default
  html_document:
    theme: flatly
    toc: true
    toc_float: true
params:
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(flextable); library(glue)
library(stepssurvey)

cfg    <- params$config
tables <- readRDS(file.path(cfg$data_path, "..", "processed", "tables.rds"))
plots  <- readRDS(file.path(cfg$data_path, "..", "processed", "plots.rds"))

show_table <- function(tbl) { if (!is.null(tbl)) tbl else cat("*Data not available*") }
```

# `r cfg$country_name` — WHO STEPS NCD Risk Factor Survey `r cfg$survey_year`

**Age range:** `r cfg$age_min`–`r cfg$age_max` years
**Survey instrument:** WHO STEPS v3.x

---

# Step 1: Behavioural Risk Factors

## Tobacco Use

```{r}
show_table(tables$current_tobacco)
```

```{r fig.width=8, fig.height=5}
if (!is.null(plots$tobacco_by_sex)) plots$tobacco_by_sex
```

## Alcohol Use

```{r}
show_table(tables$current_alcohol)
```

## Physical Activity

```{r}
show_table(tables$insufficient_pa)
```

## Diet

```{r}
show_table(tables$low_fruit_veg)
```

---

# Step 2: Physical Measurements

## Overweight and Obesity

```{r}
show_table(tables$overweight_obese)
```

```{r fig.width=10, fig.height=5}
if (!is.null(plots$obesity_by_age)) plots$obesity_by_age
```

## Blood Pressure

```{r}
show_table(tables$raised_bp)
```

```{r fig.width=10, fig.height=5}
if (!is.null(plots$bp_by_age)) plots$bp_by_age
```

---

# Step 3: Biochemical Measurements

## Blood Glucose

```{r}
show_table(tables$raised_glucose)
```

## Cholesterol

```{r}
show_table(tables$raised_chol)
```

---

*Generated by the WHO STEPS R Analysis Pipeline | `r Sys.Date()`*
